\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-24 paper/algorithm.tex
paper: cleaned up proof, first pass

>>>>

\subsection{Optimality}
We prove the optimality of the final plan selected by showing that our cost-model and algorithm
construct a complete Pareto frontier.

\begin{definition}
An abstract imputation operator $I: \mathcal{P}(String) \rightarrow \mathcal{P}(\{ \delta, \mu \})$ is a function that given a dirty set maps to
a set of imputation operations. For convenience we denote applying the set of transformations to a plan $p$ as $I(p)$ and understand that this
is shorthand for $\{ op(p)   | op \in I(\textsc{Dirty}(p)) \}$.
\end{definition}

Algorithm{~\ref{algo:top-level-planner} plans the query in stages. It first handles any pushed-down selections, it then optimizes joins, then groupby/aggregations,
and finally projections. It maintains all necessary state information through a combination of dirty sets on intermediate plans and
initializing empty plan caches at stages that require it. Given this, we can succinctly describe possible transformations to a plan at a given
point in the algorithm solely through $I$.

\begin{theorem}\label{theorem:no-bad-sit}
By computing dirty-set specific Pareto frontiers, it is never the case that we discard a plan $p_1$ s.t. $\doms{p_2}{p_1}$, but $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
\end{theorem}

Theorem~\ref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in Figure~\ref{fig:bad-sit-diagram}

\begin{figure}
\begin{tikzpicture}
% before I
\node(p1) {$p_1$};
\node[right=of p1](dom1){$\prec$};
\node[right=of dom1](p2) {$p_2$};

% after I
\node[below=of p1](p1'){$p_1'$};
\node[right=of p1'](dom2) {$\succ$};
\node[right=of dom2](p2'){$p_2'$};

% arrows
\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
\end{tikzpicture}
\caption{A bad Pareto frontier selection would choose $p_2$ and never reach $p_1'$}
\label{fig:bad-sit-diagram}
\end{figure}

\begin{proof}
For contradiction, let $p_1$ and $p_2$ be plans such that $\doms{p_2}{p_1}$, $p_1$ is removed from the plan cache
in favor of $p_2$, but there  $\exists p_1' \in I(p_1), p_2' \in I(p_2). \doms{p_1'}{p_2'} \land \textsc{Dirty}(p_1') = \textsc{Dirty}(p_2')$
and $p_1'$ is not reachable from $p_2$ by $I$.

There are two possible cases :

\begin{case}
$\textsc{Dirty}(p_1) = \textsc{Dirty}(p_2)$, in which case $I(p_1) = I(p_2)$, as $I$ uniquely determines the possible transformations. This makes $p_2$ and $p_1$ substitutable, so $p_1'$ is reachable from $p_2$. This is a contradiction.
\end{case}

\begin{case}
$\textsc{Dirty}(p_1) \not= \textsc{Dirty}(p_2)$, in which case $p_1$ is not removed from the plan cache as the $\doms{p_2}{p_1}$ is not within the same dirty set. This is a contradiction
\end{case}

So it must be the case that no such situation arises.
\end{proof}

Given that no desirable plans are discarded by computing Pareto frontiers for each dirty set, the final Pareto frontier (for dirty set $\emptyset$) is optimal.
